---
title: "Embryo RNA maternal Analysis"
author: "Leonard C. Steg"
date: '2025-07-15'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(edgeR)
  library(RColorBrewer)
  library(ggplot2)
  library(SummarizedExperiment)
  library(cowplot)
  library(SEtools)
  library(gprofiler2)
  library(rtracklayer)
  library(ggrepel)
  library(TxDb.Mmusculus.UCSC.mm10.knownGene)
  library(org.Mm.eg.db)
  library(dplyr)
})
```


Read in pheno file with metadata
```{r}
pheno <- read.csv("Steg2026/EmbryoRNA/pheno.csv", row.names = 1)
```

Load in results from DEA
```{r}
se.mat <- readRDS("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/SE.rds")

output <- "Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/"
```

Plot volcano plot
```{r}
source("Steg2026/scripts/VolcanoPlotGGPlot.R")

VolcanoPlotGGPlot(dataframe = as.data.frame(rowData(se.mat)), logFC = "dea.logFC", PValues = "dea.padj",
                  xlimits = 5, ylimits = 8.5,
                  color.highlight ="#9C44D3" ,output = output)
```

Plot GO barplot
```{r}
source("Steg2026/scripts/GoTermBarplot.R")

GoTermBarplot(GO.res = metadata(se.mat)$go_merged, color.low = "#E1E1E1", color.high = "#9C44D3",xlimits = 28,
              output = output,print = T)
```

Plot overlap between maternal and paternal transcriptome DEAs
```{r}
# Load paternal results
se.pat <- readRDS("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_pat_mother/SE.rds")

res.pat <- as.data.frame(rowData(se.pat))
res.mat <- as.data.frame(rowData(se.mat))

pat.up <- res.pat[res.pat$dea.padj < 0.05 & res.pat$dea.logFC > 0.5,]
mat.up <- res.mat[res.mat$dea.padj < 0.05 & res.mat$dea.logFC > 0.5,]

pat.down <- res.pat[res.pat$dea.padj < 0.05 & res.pat$dea.logFC <  -0.5,]
mat.down <- res.mat[res.mat$dea.padj < 0.05 & res.mat$dea.logFC < -0.5,]

overlap.up   <- pat.up[rownames(pat.up) %in% rownames(mat.up),]
overlap.down <- pat.down[rownames(pat.down) %in% rownames(mat.down),]


# Plot as barplot (counts)
df <- as.data.frame(matrix(NA, nrow = 6, ncol = 3,
                           dimnames = list(NULL, c("Direction", "Overlap", "Number"))))

df$Direction <- c(rep("Up", 3), rep("Down", 3))
df$Overlap   <- rep(c("Paternal specific", "Overlap", "Maternal specific"), 2)

df[1, "Number"] <- nrow(pat.up)     - nrow(overlap.up)
df[2, "Number"] <- nrow(overlap.up)
df[3, "Number"] <- nrow(mat.up)   - nrow(overlap.up)

df[4, "Number"] <- nrow(pat.down)   - nrow(overlap.down)
df[5, "Number"] <- nrow(overlap.down)
df[6, "Number"] <- nrow(mat.down) - nrow(overlap.down)

# Order factors: consistent stack + Down left, Up right
df$Overlap <- factor(df$Overlap, levels = c("Paternal specific", "Overlap", "Maternal specific"))
df$Direction <- factor(df$Direction, levels = c("Down", "Up"))

# Totals for label above bars
totals <- df %>%
  group_by(Direction) %>%
  summarise(Total = sum(Number), .groups = "drop")

overlap.pat.mat <- ggplot(df, aes(x = Direction, y = Number, fill = Overlap)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = ifelse(Number > 0, Number, "")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  geom_text(
    data = totals,
    aes(x = Direction, y = Total, label = Total),
    inherit.aes = FALSE,
    vjust = -0.3,
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Paternal specific"   = "#1E85D9",
      "Overlap"                   = "#666666",
      "Maternal specific" = "#9C44D3"
    )
  ) +
  labs(
    x = NULL,
    y = "Number of DEGs",
    fill = "Expression",
    title = "Overlapping DEGs between paternal and maternal transcriptomes"
  ) +
  theme_cowplot(font_size = 8) +
  coord_cartesian(clip = "off")

ggsave(
  "overlap.pat.mat.number.pdf",
  plot = overlap.pat.mat,
  device = NULL,
  path = "Steg2026/EmbryoRNA/results/03_dea/Overlap",
  width = 6, height = 5, units = "cm"
)

```

TF target enrichment analysis via Enrichr (https://maayanlab.cloud/Enrichr/). Use tables generated with DEA function to select gene names of differentially expressed genes and the background (total genome).
Download results from ChEA 2022 database enrichment analysis. Load in resulting df for plotting.
```{r}
df <- read.delim(paste0(output, "ChEA_mat_table.txt"), header = TRUE, sep = "\t", stringsAsFactors = FALSE)

df$TF <- sub(" .*", "", df$Term)

# Compute -log10(p) for y-axis
df$log_padj <- -log10(df$Adjusted.P.value)  # replace with correct column if needed

top_tfs <- df %>%
  arrange(desc(log_padj)) %>%
  slice_head(n = nrow(df[df$Adjusted.P.value < 0.05,]))  # or use more if desired

thres.fdr <- -log10(0.05)

tf_plot <- ggplot(df, aes(x = Odds.Ratio, y = log_padj)) +
  geom_hline(yintercept = thres.fdr, linewidth = 0.3, color = "grey40") +
  geom_point(color = "grey70", size = 1.5) +  
  geom_point(data = top_tfs, aes(x = Odds.Ratio, y = log_padj), color = "#9C44D3", size = 2) +  # highlight
  geom_text_repel(data = top_tfs,
                  aes(label = TF),
                  size = 2, 
                  box.padding = 0.3,
                  point.padding = 0.2,
                  max.overlaps = 15) +
  labs(x = "Odds ratio", y = expression(-log[10](adj.p-value))) +
  theme_cowplot(font_size = 8, line_size = 0.5)

ggsave("TF_enrichr_plot.pdf", plot = tf_plot, path=output, width = 4, height = 4.5, units = "cm")
```

maternal DEGs expressed in oocyte(total RNA)
```{r}
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
# Extract all the transcript names (TXNAME) as keys
k <- keys(txdb, keytype = "TXNAME")
# Create a mapping (tx2gene) from TXNAME (transcript IDs) to GENEID (UCSC Entrez IDs)
tx2gene <- AnnotationDbi::select(txdb, keys = k, columns = "GENEID", keytype = "TXNAME")
# Use org.Mm.eg.db to map UCSC Entrez IDs to Ensembl Gene IDs
ensembl_map <- mapIds(org.Mm.eg.db, 
                      keys = tx2gene$GENEID, 
                      column = "ENSEMBL", 
                      keytype = "ENTREZID", 
                      multiVals = "first")  # Takes only the first value if multiple mappings exist
# Add the Ensembl gene IDs to the tx2gene data frame
tx2gene$ENSEMBL <- ensembl_map[tx2gene$GENEID]
tx2gene_mapped <- tx2gene[!is.na(tx2gene$GENEID) & !is.na(tx2gene$ENSEMBL), ]
# Convert the ENSEMBL column from a list to a character vector
tx2gene_mapped$ENSEMBL <- sapply(tx2gene_mapped$ENSEMBL, `[`, 1)


sample.dir <- "Steg2026/Public_data/OocyteRNA/results/04_salmon_quant_SE/"
# Define the directories with Salmon quant.sf files
samples <- list.files(sample.dir)[grepl("Oocyte",list.files(sample.dir))]

# Map the sample names to the actual directory names with _SX suffix 
  files <- file.path(sample.dir, samples, "quant.sf")
names(files) <- samples

# Use tximport and extract counts
txi <- tximport(files, type = "salmon", tx2gene = tx2gene_mapped[, c("TXNAME", "ENSEMBL")])
counts <- txi$counts

storage.mode(counts) <- "integer"

rownames(counts) <- gconvert(query = gsub("\\..*","", rownames(counts)), organism = "mmusculus", target = "ENSG", filter_na = FALSE)$name


# df = raw count matrix (genes x samples)
dge.public.total <- DGEList(counts = counts)
# calculate CPM
cpm.public.total <- cpm(dge.public.total)

# filter
keep.public.total <- rowSums(cpm.public.total > 0.1) >= 2

# Select significant genes in maternal transcriptome
fdr.logfc.mat <- res.mat[res.mat$dea.padj < 0.05 & abs(res.mat$dea.logFC) > 0.5,]

# Which percentage of maternal DEGs are expressed in oocyte dataset?
perc.expressed.oocyte <- sum(rownames(fdr.logfc.mat) %in% rownames(counts[keep.public.total,]))/ nrow(fdr.logfc.mat)

perc.not.expressed.oocyte <- 1-perc.expressed.oocyte

df <- data.frame(
  category = c("Expressed in oocyte", "Not expressed in oocyte"),
  value = c(perc.expressed.oocyte, perc.not.expressed.oocyte)
)

DEGs.oocyte.plot <- ggplot(df, aes(x = "", y = value, fill = category)) +
  geom_col() +
  scale_fill_manual(values = c("Expressed in oocyte" = "#9C44D3",
                                "Not expressed in oocyte" = "#E6E6E6")) +
  theme_cowplot()

ggsave("DEGs.expressed.oocyte.pdf", plot = DEGs.oocyte.plot, path=output, width = 4, height = 4.5, units = "cm")
```

Correlation between miRNA target enrichment and maternal expression.
miRNA target enrichment was performed with EnrichR (https://maayanlab.cloud/Enrichr/). Input: Tables generated with DEA function to select gene names of differentially expressed genes and the background.
Output: miRTarBase database, results dataframe saved and loaded in.
```{r}
# Load in results from enrichR
df.mat.mirTar.up <- read.delim("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/miRTarBase_2017_up.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

df.mat.mirTar.down <- read.delim("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/miRTarBase_2017_down.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# Load in results from sperm miRNA DEA
se.miRNA <- readRDS("Steg2026/SpermSmallRNA/results/03_dea/se.miRNA.rds")
res.miRNA <- as.data.frame(rowData(se.miRNA))

# Adjust naming
df.mat.mirTar.up$miRNA.ID <- sapply(strsplit(df.mat.mirTar.up$Term, "-"), function(x) paste(x[3:length(x)], collapse = "-"))

df.mat.mirTar.down$miRNA.ID <- sapply(strsplit(df.mat.mirTar.down$Term, "-"), function(x) paste(x[3:length(x)], collapse = "-"))

res.miRNA$miRNA.ID <- sapply(strsplit(rownames(res.miRNA), "-"), function(x) paste(x[3:length(x)], collapse = "-"))

# Invert the odds ratio and combined score in the df.mat.mirTar.down
df.mat.mirTar.down$Odds.Ratio <- -(df.mat.mirTar.down$Odds.Ratio)
df.mat.mirTar.down$Combined.Score <- -(df.mat.mirTar.down$Combined.Score)

# Merge DFs
df.mat.mirTar.merged <- rbind(df.mat.mirTar.up,df.mat.mirTar.down)

# Create DF for plotting
df.cor <- left_join(df.mat.mirTar.merged[,c( "miRNA.ID", "P.value", "Odds.Ratio","Combined.Score")],
                    res.miRNA[,c("miRNA.ID","dea.logFC", "dea.PValue", "dea.padj" )])

df.cor <- df.cor[complete.cases(df.cor),]

cor.test(df.cor$Odds.Ratio, df.cor$dea.logFC)

# Define highlighted datapoints
df.cor$highlight <- with(df.cor, Odds.Ratio > 4 & dea.logFC < -0.3)

# Plot
spermmiRNA_matRNA <- ggplot(df.cor, aes(x = dea.logFC, y = Odds.Ratio)) + 
    geom_point(aes(fill = highlight),
             shape = 21, size = 2, alpha = 1,
             color = "black", stroke = 0.2) +
  # cross at (0,0)
  geom_vline(xintercept = 0, linewidth = 0.3,  color = "grey30") +
  geom_hline(yintercept = 0, linewidth = 0.3,  color = "grey30") +  
  # thin lines at x=5 and y=-0.5
  geom_vline(xintercept = -0.3, linewidth = 0.3, linetype = "dashed",color = "grey40") +
  geom_hline(yintercept = 4, linewidth = 0.3, linetype = "dashed",color = "grey40") +
  scale_fill_manual(
    values = c(`TRUE` = "#9C44D3", `FALSE` = "grey70"),
    labels = c(`TRUE` = "miRNA regulated", `FALSE` = "other"),
    name = NULL
  ) +
  labs(x = "sperm miRNA logFC", y = "miRNA Target enrichment") +
  theme_cowplot(font_size = 7) +
  xlim(-3,3) + 
  ylim(-35, 65) +
  theme(legend.position = "top")


ggsave("spermmiRNA_matRNA.correlation.pdf", plot = spermmiRNA_matRNA, path=output, width = 5.3, height = 5, units = "cm")
```

Repeat same analysis with miRNA target enrichment in paternal transcriptome
```{r}
# Load in results from enrichR
df.pat.mirTar.up <- read.delim("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_pat_mother/miRTarBase_2017_up.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

df.pat.mirTar.down <- read.delim("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_pat_mother/miRTarBase_2017_down.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# Load in results from sperm miRNA DEA
se.miRNA <- readRDS("Steg2026/SpermSmallRNA/results/03_dea/se.miRNA.rds")
res.miRNA <- as.data.frame(rowData(se.miRNA))

# Adjust naming
df.pat.mirTar.up$miRNA.ID <- sapply(strsplit(df.pat.mirTar.up$Term, "-"), function(x) paste(x[3:length(x)], collapse = "-"))

df.pat.mirTar.down$miRNA.ID <- sapply(strsplit(df.pat.mirTar.down$Term, "-"), function(x) paste(x[3:length(x)], collapse = "-"))

res.miRNA$miRNA.ID <- sapply(strsplit(rownames(res.miRNA), "-"), function(x) paste(x[3:length(x)], collapse = "-"))

# Invert the odds ratio and combined score in the df.pat.mirTar.down
df.pat.mirTar.down$Odds.Ratio <- -(df.pat.mirTar.down$Odds.Ratio)
df.pat.mirTar.down$Combined.Score <- -(df.pat.mirTar.down$Combined.Score)

# Merge DFs
df.pat.mirTar.merged <- rbind(df.pat.mirTar.up,df.pat.mirTar.down)

# Create DF for plotting
df.cor <- left_join(df.pat.mirTar.merged[,c( "miRNA.ID", "P.value", "Odds.Ratio","Combined.Score")],
                    res.miRNA[,c("miRNA.ID","dea.logFC", "dea.PValue", "dea.padj" )])

df.cor <- df.cor[complete.cases(df.cor),]

cor.test(df.cor$Odds.Ratio, df.cor$dea.logFC)

# Define highlighted datapoints
df.cor$highlight <- with(df.cor, Odds.Ratio > 4 & dea.logFC < -0.3)

# Plot
spermmiRNA_patRNA <- ggplot(df.cor, aes(x = dea.logFC, y = Odds.Ratio)) + 
    geom_point(aes(fill = highlight),
             shape = 21, size = 2, alpha = 1,
             color = "black", stroke = 0.2) +
  # cross at (0,0)
  geom_vline(xintercept = 0, linewidth = 0.3,  color = "grey30") +
  geom_hline(yintercept = 0, linewidth = 0.3,  color = "grey30") +  
  # thin lines at x=5 and y=-0.5
  geom_vline(xintercept = -0.3, linewidth = 0.3, linetype = "dashed",color = "grey40") +
  geom_hline(yintercept = 4, linewidth = 0.3, linetype = "dashed",color = "grey40") +
  scale_fill_manual(
    values = c(`TRUE` = "#1E85D9", `FALSE` = "grey70"),
    labels = c(`TRUE` = "miRNA regulated", `FALSE` = "other"),
    name = NULL
  ) +
  labs(x = "sperm miRNA logFC", y = "miRNA Target enrichment") +
  theme_cowplot(font_size = 7) +
  xlim(-3,3) + 
  ylim(-35, 65) +
  theme(legend.position = "top")


ggsave("spermmiRNA_patRNA.correlation.pdf", plot = spermmiRNA_patRNA, path=output, width = 5.3, height = 5, units = "cm")
```

Save bed files with regions for heatmaps
```{r}
# Find the Entrez ID for each maternal genes
res.mat$Entrez <- mapIds(org.Mm.eg.db,
                         keys = rownames(res.mat),
                         keytype = "SYMBOL",
                         column = "ENTREZID",
                         multiVals = "first")

# Load in txdb for mm10
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
tx_by_gene <- transcriptsBy(txdb, by = "gene")

genes <- genes(TxDb.Mmusculus.UCSC.mm10.knownGene)

set.seed(123)

# Bed of DEGs
genes.sign <- genes[names(genes) %in% unname(res.mat[res.mat$dea.padj < 0.05,]$Entrez) ]
genes.tss.sign <- resize(genes.sign, width = 1, fix = "start") 
export.bed(genes.tss.sign, con = paste0(output,"genes.tss.sign.bed"))

# Bed of random background
genes.background <- tx_by_gene[names(genes) %in% unname(res.mat$Entrez)]
genes.tss.background <- resize(genes.background, width = 1, fix = "start") 
export.bed(sample(unlist(genes.tss.background), size = nrow(res.mat[res.mat$dea.padj < 0.05,])), con = paste0(output,"genes.tss.background.bed"))
```

Heatmap for H3K4me3 over gene start sides but with all genes, leaving out up and down, including maternal tracks
```{bash engine.opts='-l', eval=F}
computeMatrix reference-point -R Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/genes.tss.sign.bed Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/genes.tss.background.bed --referencePoint "center" -S Steg2026/Public_data/results/H3K4me3/Sperm_H3K4me3_rep1.bw  Steg2026/Public_data/results/H3K4me3/PN5-zygote_H3K4me3_genome1.merged.bw Steg2026/Public_data/results/H3K4me3/2-cell-early_H3K4me3_genome1.merged.bw Steg2026/Public_data/results/H3K4me3/2-cell-late_H3K4me3_genome1.merged.bw -b 2000 -a 2000 --skipZeros -p 4 -o Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/H3K4me3.genes.tss.sign.mat.gz --outFileSortedRegions Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/H3K4me3.genes.tss.sign.bed

plotHeatmap -m Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/H3K4me3.genes.tss.sign.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 4 --heatmapWidth 2 --colorMap 'Blues' 'Purples' 'Purples' 'Purples' --whatToShow 'plot, heatmap and colorbar' --plotType se --samplesLabel "Sperm" "PN5 zygote mat" "TwoCell early mat" "TwoCell late mat" --regionsLabel "Maternal DEGs" "Random background" -out Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/H3K4me3.genes.tss.sign.pdf
```

Heatmap for ATAC DNAme and H3K27me3 in sperm cells at the maternal DEGs
```{bash engine.opts='-l', eval=F}
computeMatrix reference-point -R Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/genes.tss.sign.bed Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/genes.tss.background.bed --referencePoint "center" -S Steg2026/Other_data/SpermATAC/tracks_merged.bw Steg2026/Other_data/Sperm_EMseq/tracks_merged.bw Steg2026/Public_data/results/H3K27me3/Sperm_H3K27me3.bw -b 2000 -a 2000 --skipZeros -p 4 -o Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/Sperm.chromatin.genes.tss.sign.mat.gz --outFileSortedRegions Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/Sperm.chromatin.genes.tss.sign.bed

plotHeatmap -m Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/Sperm.chromatin.genes.tss.sign.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 4 --heatmapWidth 2 --colorMap 'Blues' --whatToShow 'plot, heatmap and colorbar' --plotType se --samplesLabel "Sperm ATAC" "Sperm DNAme" "Sperm H3K27me3" --regionsLabel "Maternal DEGs" "Random background" -out Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/Sperm.chromatin.genes.tss.sign.pdf
```