---
title: "Embryo RNA paternal Analysis"
author: "Leonard C. Steg"
date: '2025-07-15'
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(edgeR)
  library(RColorBrewer)
  library(ggplot2)
  library(SummarizedExperiment)
  library(cowplot)
  library(SEtools)
  library(gprofiler2)
  library(rtracklayer)
  library(ggrepel)
  library(TxDb.Mmusculus.UCSC.mm10.knownGene)
  library(org.Mm.eg.db)
  library(dplyr)
})
```

Read in pheno file with metadata
```{r}
pheno <- read.csv("Steg2026/EmbryoRNA/pheno.csv", row.names = 1)
```

Load in results from DEA
```{r}
se.twocell.pat <- readRDS("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_pat_mother/SE.rds")

output <- "Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_pat_mother/"
```

Plot volcano plot
```{r}
source("Steg2026/scripts/VolcanoPlotGGPlot.R")

VolcanoPlotGGPlot(dataframe = as.data.frame(rowData(se.twocell.pat)), logFC = "dea.logFC", PValues = "dea.padj",
                  xlimits = 5, ylimits = 8.5,
                  color.highlight ="#1E85D9" ,output = output)
```

Plot GO barplot
```{r}
source("Steg2026/scripts/GoTermBarplot.R")

GoTermBarplot(GO.res = metadata(se.twocell.pat)$go_merged, color.low = "#E1E1E1", color.high = "#09263E",xlimits = 28,
              output = output, print = T)
```

TF target enrichment analysis via Enrichr (https://maayanlab.cloud/Enrichr/). Use tables generated with DEA function to select gene names of differentially expressed genes and the background (total genome).
Download results from ChEA 2022 database enrichment analysis. Load in resulting df for plotting.
```{r}
df <- read.delim(paste0(output, "ChEA_pat_table.txt"), header = TRUE, sep = "\t", stringsAsFactors = FALSE)

df$TF <- sub(" .*", "", df$Term)

# Compute -log10(p) for y-axis
df$log_padj <- -log10(df$Adjusted.P.value)  # replace with correct column if needed

top_tfs <- df %>%
  arrange(desc(log_padj)) %>%
  slice_head(n = nrow(df[df$Adjusted.P.value < 0.05,]))  # or use more if desired

thres.fdr <- -log10(0.05)

tf_plot <- ggplot(df, aes(x = Odds.Ratio, y = log_padj)) +
  geom_hline(yintercept = thres.fdr, linewidth = 0.3, color = "grey40") +
  geom_point(color = "grey70", size = 1.5) +  
  geom_point(data = top_tfs, aes(x = Odds.Ratio, y = log_padj), color = "#1e85d9", size = 2) +  # highlight
  geom_text_repel(data = top_tfs,
                  aes(label = TF),
                  size = 2, 
                  box.padding = 0.3,
                  point.padding = 0.2,
                  max.overlaps = 15) +
  labs(x = "Odds ratio", y = expression(-log[10](adj.p-value))) +
  theme_cowplot(font_size = 8, line_size = 0.5)

ggsave("TF_enrichr_plot.pdf", plot = tf_plot, path=output, width = 4, height = 4.5, units = "cm")
```

Load results from analysis in spermRNA and embryoRNA maternal transcriptome
```{r}
se.sperm <- readRDS("Steg2026/SpermRNA/results/03_dea/B6-F0-Male-Basemodel_EmbryoSamples/SE.rds")

se.twocell.mat <- readRDS("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_mat_mother/SE.rds")
```

Take differentially expressed genes in paternal transcriptome and subset for genes detected in all three analyses
```{r}
res.sperm <- as.data.frame(rowData(se.sperm))
res.twocell.pat <- as.data.frame(rowData(se.twocell.pat))
res.twocell.mat <- as.data.frame(rowData(se.twocell.mat))

top.twocell.pat <- rownames(res.twocell.pat[res.twocell.pat$dea.padj < 0.05,])

top.twocell.pat <- top.twocell.pat[top.twocell.pat %in% rownames(se.sperm) & top.twocell.pat %in% rownames(se.twocell.mat)]
```

Identify genes that are consistently up or down in sperm and 2cell paternal
```{r}
logfc_matrix_sperm2pat <- data.frame(
  Sperm = res.sperm[top.twocell.pat, "dea.logFC"],
  TwoCell = res.twocell.pat[top.twocell.pat, "dea.logFC"],
  row.names = top.twocell.pat
)
# Subset for genes with all positive logFCs
logfc_up <- logfc_matrix_sperm2pat[apply(logfc_matrix_sperm2pat, 1, function(x) all(x > 0.5)), ]

# Subset for genes with all negative logFCs
logfc_down <- logfc_matrix_sperm2pat[apply(logfc_matrix_sperm2pat, 1, function(x) all(x < -0.5)), ]

Sperm2EmbryoGenes <- rbind(logfc_up, logfc_down)

write.csv(Sperm2EmbryoGenes, file = "Steg2026/EmbryoRNA/results/04_sperm_to_embryo/Sperm2EmbryoGenes.csv")
```

Plot heatmap with expression of Sperm2EmbryoGenes in sperm, embryo paternal and maternal transcriptome.
```{r}
range_val <- max(abs(logfc_matrix_sperm2pat)) 

logfc_mat <- data.frame(
  Sperm = res.sperm[top.twocell.pat, "dea.logFC"],
  TwoCell.pat = res.twocell.pat[top.twocell.pat, "dea.logFC"],
  TwoCell.mat = res.twocell.mat[top.twocell.pat, "dea.logFC"],
  row.names = top.twocell.pat
)

logfc_mat <- logfc_mat[rownames(Sperm2EmbryoGenes),]

# Plot
sperm2embryo.heatmap <- pheatmap(logfc_mat, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         show_rownames = F,
         fontsize = 7,
         scale = "column",           # or "row" to z-score per gene
         color = colorRampPalette(c("#761111", "white", "#114476"))(100),
         breaks = seq(-range_val, range_val, length.out = 101),
         main = "logFC heatmap of 2cell DEA top genes")


pdf(file = "Steg2026/EmbryoRNA/results/04_sperm_to_embryo/Sperm2EmbryoHeatmap.pdf", width = 2.5,height = 2.5)
sperm2embryo.heatmap
dev.off()
```


TF target enrichment analysis via Enrichr (https://maayanlab.cloud/Enrichr/). Use Sperm2Embryo genes as input and total transcriptome genes as background.
Download results from ChEA 2022 database enrichment analysis. Load in resulting df for plotting.
```{r}
df <- read.delim("Steg2026/EmbryoRNA/results/04_sperm_to_embryo/ChEA_2022_merged.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

df$TF <- sub(" .*", "", df$Term)

# Compute -log10(p) for y-axis
df$log_padj <- -log10(df$Adjusted.P.value)  # replace with correct column if needed

top_tfs <- df %>%
  arrange(desc(log_padj)) %>%
  slice_head(n = nrow(df[df$Adjusted.P.value < 0.05,]))  # or use more if desired

tf_plot <- ggplot(df, aes(x = Odds.Ratio, y = log_padj)) +
  geom_hline(yintercept = thres.fdr, linewidth = 0.3, color = "grey40") +
  geom_point(color = "grey70", size = 1.5) +  
  geom_point(data = top_tfs, aes(x = Odds.Ratio, y = log_padj), color = "#5A7B9C", size = 2) +  # highlight
  geom_text_repel(data = top_tfs,
                  aes(label = TF),
                  size = 2, 
                  box.padding = 0.3,
                  point.padding = 0.2,
                  max.overlaps = 15) +
  labs(x = "Odds ratio", y = expression(-log[10](adj.p-value))) +
  theme_cowplot(font_size = 8, line_size = 0.5)

ggsave("Sperm2Embryo.TF_enrichr_plot.pdf", plot = tf_plot, path="Steg2026/EmbryoRNA/results/04_sperm_to_embryo/", width = 4, height = 4.5, units = "cm")

```

Analysis to find overlapping most expressed transcripts of Sperm2Embryo genes in both sperm and paternal transcriptome of twocell embryo
Step 1: Create count matrix of transcripts in both datasets

```{bash engine.opts='-l', eval=F}
python Steg2026/scripts/CreateCounts1.py Steg2026/EmbryoRNA/results/01_preprocess/15_salmon_post_snpsplit/genome1/*_quant/quant.sf Steg2026/EmbryoRNA/results/01_preprocess/15_salmon_post_snpsplit/genome1/count_matrix.tsv

python Steg2026/scripts/CreateCounts1.py Steg2026/SpermRNA/results/01_preprocess/12_salmon_post_dedup/*_quant/quant.sf Steg2026/SpermRNA/results/01_preprocess/12_salmon_post_dedup/count_matrix.tsv
```

Load tables and annotation, find most expressed transcripts per gene in both datasets, calculate proportion and plot.
```{r}
# Load in Twocell paternal transcript count matrix
counts.trx.pat <- read.table(
  "Steg2026/EmbryoRNA/results/01_preprocess/15_salmon_post_snpsplit/genome1/count_matrix.tsv",
  header = TRUE,
  sep = "\t",
  row.names = 1,
  check.names = FALSE
)

# Load in sperm transcript count matrix
counts.trx.sperm <- read.table(
  "Steg2026/SpermRNA/results/01_preprocess/12_salmon_post_dedup/count_matrix.tsv",
  header = TRUE,
  sep = "\t",
  row.names = 1,
  check.names = FALSE
)

# Subset for samples of fathers used for IVF
samples.NOT.embryo.fathers <- c("B6-F0-03", "B6-F0-06", "B6-F0-09", "B6-F0-20", "B6-F0-25", "B6-F0-38", "B6-F0-61", "B6-F0-64")
counts.trx.sperm <- counts.trx.sperm[,!(colnames(counts.trx.sperm) %in% samples.NOT.embryo.fathers)]

# Load annotation files
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
# Extract all the transcript names (TXNAME) as keys
k <- keys(txdb, keytype = "TXNAME")
# Create a mapping (tx2gene) from TXNAME (transcript IDs) to GENEID (UCSC Entrez IDs)
tx2gene <- select(txdb, keys = k, columns = "GENEID", keytype = "TXNAME")
# Use org.Mm.eg.db to map UCSC Entrez IDs to Ensembl Gene IDs
ensembl_map <- mapIds(org.Mm.eg.db, 
                      keys = tx2gene$GENEID, 
                      column = "SYMBOL", 
                      keytype = "ENTREZID", 
                      multiVals = "first") 

tx2gene$NAME <- as.character(ensembl_map[tx2gene$GENEID])
tx2gene <- tx2gene[!is.na(tx2gene$GENEID),]

# Create dataframe with most expressed transcripts in each dataset per Sperm2Embryo gene
colnames.df <- c("SpermMostUsed", "%Sperm", "EmbryoMostUsed", "%Embryo", "Overlap", "InTop3", "%inTop3", "InTop3AndHigh")

Sperm2Embryo.trxs <- as.data.frame(matrix(NA, nrow = nrow(Sperm2EmbryoGenes), 
                                          ncol = length(colnames.df), 
                                          dimnames = list(rownames(Sperm2EmbryoGenes), colnames.df)))

for (i in 1:nrow(Sperm2EmbryoGenes)) {
  
  gene <- rownames(Sperm2EmbryoGenes)[i]
  
  trxs <- tx2gene[tx2gene$NAME == gene,]$TXNAME
  trxs <- trxs[!is.na(trxs)]
  
  if (length(trxs > 1)) {
    sperm.sub <- counts.trx.sperm[trxs,]
    sperm.sub.sums <- rowSums(sperm.sub)
    sperm.sub.sums <- sperm.sub.sums[!is.na(sperm.sub.sums)]
    sperm.most.trx <- names(sperm.sub.sums[sperm.sub.sums == max(sperm.sub.sums)])
    sperm.top3.trx <- names(sperm.sub.sums[order(sperm.sub.sums, decreasing = T)][1:3])
    
    embryo.sub <- counts.trx.pat[trxs,]
    embryo.sub.sums <- rowSums(embryo.sub)
    embryo.sub.sums <- embryo.sub.sums[!is.na(embryo.sub.sums)]
    embryo.most.trx <- names(embryo.sub.sums[embryo.sub.sums == max(embryo.sub.sums)])
    
    Sperm2Embryo.trxs[i,]$SpermMostUsed <- sperm.most.trx
    Sperm2Embryo.trxs[i,]$`%Sperm` <- max(sperm.sub.sums)/sum(sperm.sub.sums)
    
    Sperm2Embryo.trxs[i,]$EmbryoMostUsed <- embryo.most.trx
    Sperm2Embryo.trxs[i,]$`%Embryo` <- max(embryo.sub.sums)/sum(embryo.sub.sums)
    
    Sperm2Embryo.trxs[i,]$Overlap <- embryo.most.trx == sperm.most.trx
    Sperm2Embryo.trxs[i,]$InTop3 <- embryo.most.trx %in% sperm.top3.trx
    Sperm2Embryo.trxs[i,]$`%inTop3` <- sperm.sub.sums[embryo.most.trx]/sum(sperm.sub.sums)
    
    Sperm2Embryo.trxs[i,]$InTop3AndHigh <- Sperm2Embryo.trxs[i,]$InTop3 == T & Sperm2Embryo.trxs[i,]$`%inTop3` > 0.4
    
  }
}

Sperm2Embryo.trxs <- Sperm2Embryo.trxs[complete.cases(Sperm2Embryo.trxs),]

# Proportion of genes that have the same transcript
prop.sperm2embryo.trxs.overlap <- sum(Sperm2Embryo.trxs$Overlap)/nrow(Sperm2Embryo.trxs)
# 0.5875486

# Plot barplot
df <- tibble(
  category = c("Shared most expressed transcript", "Different transcript"),
  proportion = c(
    prop.sperm2embryo.trxs.overlap,
    1 - prop.sperm2embryo.trxs.overlap
  )
)

barplot.trxs <- ggplot(df, aes(x = "", y = proportion, fill = category)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(
    values = c(
      "Shared most expressed transcript" = "#5A7B9C",
      "Different transcript" = "#B3B3B3"
    )
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    x = NULL,
    y = "Percentage of genes",
    fill = NULL
  ) +
  theme_cowplot(font_size = 8)


ggsave(filename = "barplot.overlap.trxs.pdf", plot = barplot.trxs,path = "Steg2026/EmbryoRNA/results/04_sperm_to_embryo/", width = 7, height = 4, units = "cm")
```


Create .bed files at TSS of Sperm2Embryo Genes
```{r}
# Find the Entrez ID for each of the sperm to embryo genes
Sperm2EmbryoGenes$Entrez <- mapIds(org.Mm.eg.db,
                                           keys = rownames(Sperm2EmbryoGenes),
                                           keytype = "SYMBOL",
                                           column = "ENTREZID",
                                           multiVals = "first")

# Load in txdb for mm10
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
tx_by_gene <- transcriptsBy(txdb, by = "gene")

genes <- genes(TxDb.Mmusculus.UCSC.mm10.knownGene)

# Prepare output and seed
output <- "Steg2026/EmbryoRNA/results/04_sperm_to_embryo/"

set.seed(123)

# Bed for Sperm to Embryo genes
genes.merged <- genes[names(genes) %in% unname(Sperm2EmbryoGenes$Entrez) ]
genes.tss.merged <- resize(genes.merged, width = 1, fix = "start") 
export.bed(genes.tss.merged, con = paste0(output,"genes.tss.merged.bed"))

# Bed for random background take genes expressed in early 2 cell embryo
SE.embryo<- readRDS("Steg2026/EmbryoRNA/results/03_dea/TwoCell-Basemodel_dedup_mother/SE.rds")
res.embryo <- as.data.frame(rowData(SE.embryo))

res.embryo$Entrez <- mapIds(org.Mm.eg.db,keys = rownames(res.embryo),keytype = "SYMBOL", column = "ENTREZID",multiVals = "first")

genes.background <- tx_by_gene[names(genes) %in% unname(res.embryo$Entrez)]
genes.tss.background <- resize(genes.background, width = 1, fix = "start") 
export.bed(sample(unlist(genes.tss.background), size = nrow(Sperm2EmbryoGenes)), con = paste0(output,"genes.tss.background.bed"))
```

Heatmap for H3K4me3 over gene start sides but with all genes, leaving out up and down, including maternal tracks
```{bash engine.opts='-l', eval=F}
computeMatrix reference-point -R Steg2026/EmbryoRNA/results/04_sperm_to_embryo/genes.tss.merged.bed Steg2026/EmbryoRNA/results/04_sperm_to_embryo/genes.tss.background.bed --referencePoint "center" -S Steg2026/Public_data/results/H3K4me3/Sperm_H3K4me3_rep1.bw Steg2026/Public_data/results/H3K4me3/PN5-zygote_H3K4me3_genome2.merged.bw Steg2026/Public_data/results/H3K4me3/2-cell-early_H3K4me3_genome2.merged.bw Steg2026/Public_data/results/H3K4me3/2-cell-late_H3K4me3_genome2.merged.bw Steg2026/Public_data/results/H3K4me3/Oocyte_H3K4me3_merged.bw Steg2026/Public_data/results/H3K4me3/PN5-zygote_H3K4me3_genome1.merged.bw Steg2026/Public_data/results/H3K4me3/2-cell-early_H3K4me3_genome1.merged.bw Steg2026/Public_data/results/H3K4me3/2-cell-late_H3K4me3_genome1.merged.bw -b 2000 -a 2000 --skipZeros -p 4 -o Steg2026/EmbryoRNA/results/04_sperm_to_embryo/H3K4me3.genes.tss.onlymerged.withMat.mat.gz --outFileSortedRegions Steg2026/EmbryoRNA/results/04_sperm_to_embryo/H3K4me3.genes.tss.onlymerged.withMat.bed

plotHeatmap -m Steg2026/EmbryoRNA/results/04_sperm_to_embryo/H3K4me3.genes.tss.onlymerged.withMat.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 4 --heatmapWidth 2 --colorMap 'Blues' 'Blues' 'Blues' 'Blues' 'Purples' 'Purples' 'Purples' 'Purples' --whatToShow 'plot, heatmap and colorbar' --plotType se --samplesLabel "Sperm" "PN5 zygote pat" "TwoCell early pat" "TwoCell late pat" "Oocyte" "PN5 zygote mat" "TwoCell early mat" "TwoCell late mat" --regionsLabel "Sperm2Embryo genes" "Random background" -out Steg2026/EmbryoRNA/results/04_sperm_to_embryo/H3K4me3.genes.tss.heatmap.onlymerged.withMat.pdf
```

Heatmap for ATAC signal in sperm
```{bash engine.opts='-l', eval=F}
computeMatrix reference-point -R Steg2026/EmbryoRNA/results/04_sperm_to_embryo/genes.tss.merged.bed Steg2026/EmbryoRNA/results/04_sperm_to_embryo/genes.tss.background.bed --referencePoint "center" -S Steg2026/Other_data/SpermATAC/tracks_merged.bw -b 2000 -a 2000 --skipZeros -p 4 -o Steg2026/EmbryoRNA/results/04_sperm_to_embryo/ATAC.sperm.mat.gz --outFileSortedRegions Steg2026/EmbryoRNA/results/04_sperm_to_embryo/ATAC.sperm.bed

plotHeatmap -m Steg2026/EmbryoRNA/results/04_sperm_to_embryo/ATAC.sperm.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 4 --heatmapWidth 2 --colorMap 'Blues' --whatToShow 'plot, heatmap and colorbar' --plotType se --samplesLabel "Sperm ATAC" --regionsLabel "Sperm2Embryo genes" "Random background" -out Steg2026/EmbryoRNA/results/04_sperm_to_embryo/ATAC.sperm.pdf
```

Heatmap for DNAme signal in sperm
```{bash engine.opts='-l', eval=F}
computeMatrix reference-point -R Steg2026/EmbryoRNA/results/04_sperm_to_embryo/genes.tss.merged.bed Steg2026/EmbryoRNA/results/04_sperm_to_embryo/genes.tss.background.bed --referencePoint "center" -S Steg2026/Other_data/Sperm_EMseq/tracks_merged.bw -b 2000 -a 2000 --skipZeros -p 8 -o Steg2026/EmbryoRNA/results/04_sperm_to_embryo/Sperm.DNAme.mat.gz --outFileSortedRegions Steg2026/EmbryoRNA/results/04_sperm_to_embryo/Sperm.DNAme.bed

plotHeatmap -m Steg2026/EmbryoRNA/results/04_sperm_to_embryo/Sperm.DNAme.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 10 --colorMap 'Greys' --whatToShow 'plot, heatmap and colorbar' --samplesLabel "Sperm DNAme" -out Steg2026/EmbryoRNA/results/04_sperm_to_embryo/Sperm.DNAme.pdf
```

Heatmap with CG content around Sperm2Embryo genes
```{bash engine.opts='-l', eval=F}
# Calcuate CG content in 10 bp windows across the genome
bedtools makewindows -g mm10.chrom.sizes -w 10 \
| bedtools nuc -fi Steg2026/references/genomes/mm10_raw/mm10.fa -bed - \
| awk '{ if (NF<5) next; cg=$5; if (cg>1) cg/=100.0; printf "%s\t%d\t%d\t%.6f\n",$1,$2,$3,cg }' \
| sort -k1,1 -k2,2n -T ./ \
> mm10_GC.bedGraph

# Convert .bedGraph file to .bw file
TMPDIR=./ bedGraphToBigWig Steg2026/references/mm10_GC.bedGraph Steg2026/references/mm10.chrom.sizes Steg2026/references/mm10_GC.bw

# Plot
computeMatrix reference-point -R Steg2026/EmbryoRNA/results/04_sperm_to_embryo/genes.tss.merged.bed Steg2026/EmbryoRNA/results/04_sperm_to_embryo/genes.tss.background.bed --referencePoint "center" -S Steg2026/references/mm10_CG/mm10_GC.bw -b 2000 -a 2000 --skipZeros -p 8 -o Steg2026/EmbryoRNA/results/04_sperm_to_embryo/CGcontent.genes.tss.mat.gz --outFileSortedRegions Steg2026/EmbryoRNA/results/04_sperm_to_embryo/CGcontent.genes.tss.bed

plotHeatmap -m Steg2026/EmbryoRNA/results/04_sperm_to_embryo/CGcontent.genes.tss.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 4 --heatmapWidth 2 --colorMap 'YlOrBr' --whatToShow 'plot, heatmap and colorbar' -min 0 -max 1 --samplesLabel "CG content" -out Steg2026/EmbryoRNA/results/04_sperm_to_embryo/CGcontent.genes.tss.pdf
```

