---
title: "ATAC DEA B6 F0 Males"
author: "Leonard C. Steg"
date: '2024-04-29'
output: html_document
---

```{r, warning=F}
suppressPackageStartupMessages({
  library(DiffBind)
  library(dplyr)
  library(rtracklayer)
  library(Rsamtools)
  library(cowplot)
  library(readxl)
  library(ggplot2)
  library(jsonlite)
  library(reshape2)
  library(edgeR)
  library(DESeq2)
  library(csaw)
  library(ChIPseeker)
  library(ChIPpeakAnno)
})
```


Load in pheno file with metadata
```{r}
pheno <- data.frame(read_csv("Steg2026/LiverATAC/pheno.csv"))
```

Define subset
```{r}
strain <- "B6"
generation <- "F0"
sex <- "Male"

output <- "LiverATAC/results/dea_03/B6-F0-Male_Basemodel/"

if (!file.exists(paste0(outputdir))){
  dir.create(paste0(outputdir))
}
```

Load DEA function
```{r}
source("LiverATAC/scripts/ATAC.dea.R")
```

Run DEA function
```{r}
ATAC.dea(pheno = pheno, bamdir = "LiverATAC/results/preprocess_01/08_NF/", peakdir = "LiverATAC/results/preprocess_01/09_peaks/",
         strain = "B6", generation = "F0", sex = "Male", 
         VarExp = "Group", CtrGroup = "CTR", ExpGroup = "LPD", 
         CovarBio = NULL, CovarTec = NULL, minOverlap = 0.3,
         outputdir = output, suffix = "Basemodel")
```

Define cutoffs
```{r}
pthreshold <- 0.05
logthreshold <- 1
```


Read in DEA results and save bed files with ranges of DARs
```{r}
# Load SE with results
SE <- readRDS("LiverATAC/results/dea_04/B6-F0-Male-Basemodel/SE.rds")

up <- SE[rowData(SE)$dea$padj < pthreshold & rowData(SE)$dea$logFC > logthreshold]
up.ranges <- granges(up)

up.ranges.df <- data.frame(seqnames=seqnames(up.ranges),
  starts=start(up.ranges)-1,
  ends=end(up.ranges),
  strands=strand(up.ranges))

write.table(up.ranges.df, file="LiverATAC/results/dea_04/B6-F0-Male-Basemodel/up.bed", quote=F, sep="\t", row.names=F, col.names=F)

down <- SE[rowData(SE)$dea$padj < pthreshold & rowData(SE)$dea$logFC < -logthreshold]
down.ranges <- granges(down)

down.ranges.df <- data.frame(seqnames=seqnames(down.ranges),
  starts=start(down.ranges)-1,
  ends=end(down.ranges),
  strands=strand(down.ranges))

write.table(down.ranges.df, file="LiverATAC/results/dea_04/B6-F0-Male-Basemodel/down.bed", quote=F, sep="\t", row.names=F, col.names=F)

# Merge bam files
ctr <- colData(SE)$Sample[colData(SE)$Group == "CTR"]
lpd <- colData(SE)$Sample[colData(SE)$Group == "LPD"]

ctrBam <- colData(SE)$bamReads[colData(SE)$Group == "CTR"]
lpdBam <- colData(SE)$bamReads[colData(SE)$Group == "LPD"]

system(paste0("samtools merge -@ 8 ", output, "ctrMerged.bam ", paste0(ctrBam, collapse = " ")))
system(paste0("samtools index ", output, "ctrMerged.bam"))
system(paste0("samtools merge -@ 8 ", output, "lpdMerged.bam ", paste0(lpdBam, collapse = " ")))
system(paste0("samtools index ", output, "lpdMerged.bam"))
```

Plot heatmaps at DARs
```{bash engine.opts='-l', eval=F}
scripts/tracks.sh LiverATAC/results/dea_04/B6-F0-Male-Basemodel/ctrMerged.bam LiverATAC/results/dea_04/B6-F0-Male-Basemodel/ logs/26_mergedtracks/ 5 8
scripts/tracks.sh LiverATAC/results/dea_04/B6-F0-Male-Basemodel/lpdMerged.bam LiverATAC/results/dea_04/B6-F0-Male-Basemodel/ logs/26_mergedtracks/ 5 8

computeMatrix reference-point -R LiverATAC/results/dea_04/B6-F0-Male-Basemodel/up.bed LiverATAC/results/dea_04/B6-F0-Male-Basemodel/down.bed --referencePoint "center" -S LiverATAC/results/dea_04/B6-F0-Male-Basemodel/ctrMerged.bw LiverATAC/results/dea_04/B6-F0-Male-Basemodel/lpdMerged.bw -b 2000 -a 2000 --skipZeros -p 8 -o LiverATAC/results/dea_04/B6-F0-Male-Basemodel/DAR.mat.gz --outFileSortedRegions LiverATAC/results/dea_04/B6-F0-Male-Basemodel/DAR.bed

plotHeatmap -m LiverATAC/results/dea_04/B6-F0-Male-Basemodel/DAR.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 4 --heatmapWidth 2 --colorMap 'Greys' --whatToShow 'plot, heatmap and colorbar' --plotType se -out LiverATAC/results/dea_04/B6-F0-Male-Basemodel/DAR.heatmap.pdf
```


Supplements

Motif analysis B6 F0 Males using Homer findMotifsGenome.pl
From the results, the enrichment scores and pvalues are manually entered into an excel sheet which is then used for plotting
```{bash engine.opts='-l', eval=F}
bed2pos.pl LiverATAC/results/dea_04/B6-F0-Male-Basemodel/up.bed > LiverATAC/results/dea_04/B6-F0-Male-Basemodel/up.pos

findMotifsGenome.pl LiverATAC/results/dea_04/B6-F0-Male-Basemodel/up.pos mm10 LiverATAC/results/dea_04/B6-F0-Male-Basemodel/up -preparsedDir LiverATAC/results/dea_04/B6-F0-Male-Basemodel/preparsed/ -size 200 -mask -p 8

bed2pos.pl LiverATAC/results/dea_04/B6-F0-Male-Basemodel/down.bed > LiverATAC/results/dea_04/B6-F0-Male-Basemodel/down.pos

findMotifsGenome.pl LiverATAC/results/dea_04/B6-F0-Male-Basemodel/down.pos mm10 LiverATAC/results/dea_04/B6-F0-Male-Basemodel/down -preparsedDir LiverATAC/results/dea_04/B6-F0-Male-Basemodel/preparsed/ -size 200 -mask -p 8
```

Load in excel sheet with enrichment scores and pvalues and plot heatmap
```{r}
enrichment <- as.data.frame(read_excel("LiverATAC/results/TFs.motifs.xlsx", sheet = 4))
rownames(enrichment) <- enrichment[,1]
enrichment <- enrichment[,-1]

target <- as.data.frame(read_excel("LiverATAC/results/TFs.motifs.xlsx", sheet = 5))
rownames(target) <- target[,1]
target <- target[,-1]


# Convert both dataframes to long (tidy) format
enrich_long <- melt(as.matrix(enrichment), varnames = c("Sample", "TF"), value.name = "Enrichment")
target_long <- melt(as.matrix(target), varnames = c("Sample", "TF"), value.name = "Target")

# Merge them to align cell by cell
heatmap_df <- merge(enrich_long, target_long, by = c("Sample", "TF"))
heatmap_df$TF <- factor(heatmap_df$TF, levels = rev(colnames(enrichment)))
# Plot with ggplot
tf_motif_heatmap <- ggplot(heatmap_df, aes(x = Sample, y = TF, fill = Enrichment)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Target, 2)), size = 2) +
  scale_fill_gradient(
    low = "white",
    high = "#7638A7",
    limits = c(0, max(heatmap_df$Enrichment, na.rm = TRUE)), 
    name = "Enrichment"
  )+
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(x = NULL, y = NULL, title = "Enrichment Heatmap with Target Values") +
  theme_cowplot(font_size = 8)

ggsave(filename =  "LiverATAC/results/F0.motif.heatmap.pdf", tf_motif_heatmap,  width = 8, height = 5, units = "cm")
```



Annotate peaks with ChipSeeker
```{r}
#Load in TxDB and function to clean up annotation dataframe
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
cleanupanno <- function(annostat) {

  rownames(annostat) <- annostat$Feature
  annostat$Feature <- NULL
  annostat <- as.data.frame(t(annostat))
  annostat$Exon <- sum(annostat$`1st Exon`, annostat$`Other Exon`)
  annostat$Intron <- sum(annostat$`1st Intron`, annostat$`Other Intron`)

  annostat$`1st Intron` <- NULL
  annostat$`Other Intron` <- NULL
  annostat$`1st Exon` <- NULL
  annostat$`Other Intron` <- NULL

  annostat <- as.data.frame(t(annostat))
  annostat$Feature <- rownames(annostat)
  
  annostat$FrequencyPercent <- annostat$Frequency/100
  
  return(annostat)
  
}

# Get everything for Down DARs
anno.down <- annotatePeak(down.ranges, TxDb = txdb,genomicAnnotationPriority = c("Promoter", "5UTR", "3UTR", "Exon", "Intron","Intergenic" ),
                     tssRegion = c(-1000, 1000))
annostat.down <- as.data.frame(anno.down@annoStat)
annostat.down <- cleanupanno(annostat = annostat.down)

annostat.down$DARs <- "Down DARs"

# Get everything for Up DARs
anno.up <- annotatePeak(up.ranges, TxDb = txdb,genomicAnnotationPriority = c("Promoter", "5UTR", "3UTR", "Exon", "Intron","Intergenic" ),
                     tssRegion = c(-1000, 1000))
annostat.up <- as.data.frame(anno.up@annoStat)
annostat.up <- cleanupanno(annostat = annostat.up)

annostat.up$DARs <- "Up DARs"

annostat.merged <- rbind(annostat.up, annostat.down)

plot <- ggplot(annostat.merged, aes(fill=Feature, y=Frequency, DARs)) + 
  geom_bar(position="fill", stat="identity") +
  scale_color_viridis(discrete=TRUE, option="cividis") + 
  theme_classic()

ggsave(paste0(output, "DAR.genomic.distribution.pdf"), plot, width = 6, height = 6)
```

Plot histone PTMs from public datasets at DARs
```{bash engine.opts='-l', eval=F}
computeMatrix reference-point -R LiverATAC/results/dea_04/B6-F0-Male-Basemodel/up.bed LiverATAC/results/dea_04/B6-F0-Male-Basemodel/down.bed --referencePoint "center" -S Public_data/Liver/08_tracks/Liver_H3K4me1.merged.bw Public_data/Liver/08_tracks/Liver_H3K4me3.merged.bw Public_data/Liver/08_tracks/Liver_H3K27ac.merged.bw -b 2000 -a 2000 --skipZeros -p 8 -o LiverATAC/results/dea_04/B6-F0-Male-Basemodel/DAR.PTMs.mat.gz --outFileSortedRegions LiverATAC/results/dea_04/B6-F0-Male-Basemodel/DAR.PTMs.bed

plotHeatmap -m LiverATAC/results/dea_04/B6-F0-Male-Basemodel/DAR.PTMs.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 4 --heatmapWidth 2 --colorMap 'GnBu_r' 'Greens_r' 'hot_r' --whatToShow 'plot, heatmap and colorbar' --plotType se --regionsLabel "H3K4me1" "H3K4me3" "H3K27ac" -out LiverATAC/results/dea_04/B6-F0-Male-Basemodel/DAR.PTMs.heatmap.pdf
```
