---
title: "LPD1 B6 F1 Female ATAC DEA"
author: "Leonard C. Steg"
date: '2024-11-04'
output: html_document
---

```{r, warning=F}
suppressPackageStartupMessages({
  library(DiffBind)
  library(dplyr)
  library(rtracklayer)
  library(Rsamtools)
  library(cowplot)
  library(readxl)
  library(ggplot2)
  library(jsonlite)
  library(reshape2)
  library(edgeR)
  library(DESeq2)
  library(csaw)
  library(ChIPseeker)
  library(ChIPpeakAnno)
})
```


Load in pheno file with metadata
```{r}
pheno <- data.frame(read_csv("Steg2026/LiverATAC/pheno.csv"))
```

Define subset
```{r}
strain <- "B6"
generation <- "F1"
sex <- "Female"

pheno <- pheno[pheno$Strain == strain & pheno$Generation == generation & pheno$Sex == sex,]

```

Create directories
```{r}
if (!file.exists("results/preprocess_01/17_fatherMerged/Females")){
   dir.create("results/preprocess_01/17_fatherMerged/Females")
}
if (!file.exists("results/preprocess_01/17_fatherMerged/Females/01_NF")){
   dir.create("results/preprocess_01/17_fatherMerged/Females/01_NF")
}
if (!file.exists("results/preprocess_01/17_fatherMerged/Females/02_downsampled")){
   dir.create("results/preprocess_01/17_fatherMerged/Females/02_downsampled")
}
if (!file.exists("results/preprocess_01/17_fatherMerged/Females/03_peaks")){
   dir.create("results/preprocess_01/17_fatherMerged/Females/03_peaks")
}
```


Merge bam files per father
```{r}
fathers <- unique(pheno$Father)

bamdir <- "results/preprocess_01/08_NF/"

bamdirMerged <- "LiverATAC/results/preprocess_01/17_fatherMerged/Females/01_NF/"

for (i in 1:length(fathers)){
  
  bam <- paste0(bamdir,pheno[pheno$Father == fathers[i],]$Sample, ".bam")
  
  system(paste0("samtools merge -@ 8 ", bamdirMerged, paste0(fathers[i], ".bam "), paste0(bam, collapse = " ")))
  system(paste0("samtools index ", bamdirMerged, paste0(fathers[i], ".bam ")))
  
}
```

Count reads per merged sample
```{bash engine.opts='-l', eval=F}
for bam_file in LiverATAC/results/preprocess_01/17_fatherMerged/Females/01_NF/*.bam; do
read_count=$(samtools view -c "$bam_file") 
echo "$bam_file: $read_count" 
done
```

Downsample all samples to lowest count and index downsampled bams
```{bash engine.opts='-l', eval=F}

for FILE in LiverATAC/results/preprocess_01/17_fatherMerged/Females/01_NF/*.bam; do scripts/downsample.sh $FILE LiverATAC/results/preprocess_01/17_fatherMerged/Females/02_downsampled/ 15000000; done

for FILE in LiverATAC/results/preprocess_01/17_fatherMerged/Females/02_downsampled/*.bam; do samtools index -@ 10 $FILE; done
```

Count reads per merged and downsampled sample
```{bash engine.opts='-l', eval=F}
for bam_file in LiverATAC/results/preprocess_01/17_fatherMerged/Females/02_downsampled/*.bam; do
read_count=$(samtools view -c "$bam_file") 
echo "$bam_file: $read_count" 
done
```

Call peaks
```{bash engine.opts='-l', eval=F}
for FILE in LiverATAC/results/preprocess_01/17_fatherMerged/Females/02_downsampled/*.bam; do LiverATAC/scripts/peakcalling.sh $FILE LiverATAC/results/preprocess_01/17_fatherMerged/Females/03_peaks/ logs/27_peaksMergedFathers/; done
```

Merge pheno for fathers
```{r}
pheno.father <- as.data.frame(matrix(NA, nrow = length(fathers), ncol = 3, dimnames = list(fathers, c("Sample", "Group", "nOffspring"))))

pheno.father$Sample <- rownames(pheno.father)
for (i in 1:nrow(pheno.father)) {
  pheno.father$Group[i] <- pheno[pheno[,"Father"] == fathers[i], "Group"][1]
}

OffspringPerFather <- table(pheno$Father)
for (i in 1:nrow(pheno.father)) {
  pheno.father$nOffspring[i] <- OffspringPerFather[rownames(pheno.father)[i]]
}

pheno.father$Strain <- strain
pheno.father$Generation <- generation
pheno.father$Sex <- sex
```

Load DEA function
```{r}
source("LiverATAC/scripts/ATAC.dea.R")
```

Run DEA function on downsampled samples
```{r}

ATAC.dea(pheno = pheno.father, bamdir = "results/preprocess_01/17_fatherMerged/Females/02_downsampled/", 
         peakdir = "results/preprocess_01/17_fatherMerged/Females/03_peaks/",
         strain = "B6", generation = "F1", sex = "Female", exclude = NULL,
         VarExp = "Group", CtrGroup = "CTR", ExpGroup = "LPD", 
         CovarBio = NULL, CovarTec = NULL, VarNested = NULL, PCAvariables = PCAvariables, minOverlap = 0.3,
         outputdir = "results/dea_merged_07/", suffix = "Basemodel_downsampled")
```
Define cutoffs and variables
```{r}
# Write bed files for up and down regulated peaks
pthreshold <- 0.05
logthreshold <- 1
output <- "LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/"
```

Read in DEA results and save bed files with ranges of DARs
```{r}
SE <- readRDS("LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/SE.rds")

# Merge bam files
ctr <- colData(SE)$Sample[colData(SE)$Group == "CTR"]
lpd <- colData(SE)$Sample[colData(SE)$Group == "LPD"]

ctrBam <- colData(SE)$bamReads[colData(SE)$Group == "CTR"]
lpdBam <- colData(SE)$bamReads[colData(SE)$Group == "LPD"]

system(paste0("samtools merge -@ 8 ", output, "ctrMerged.bam ", paste0(ctrBam, collapse = " ")))
system(paste0("samtools index ", output, "ctrMerged.bam"))
system(paste0("samtools merge -@ 8 ", output, "lpdMerged.bam ", paste0(lpdBam, collapse = " ")))
system(paste0("samtools index ", output, "lpdMerged.bam"))

# Subset for up DARs and write bed file
up <- SE[rowData(SE)$dea$PValue < pthreshold & rowData(SE)$dea$logFC > logthreshold]
up.ranges <- granges(up)

up.ranges.df <- data.frame(seqnames=seqnames(up.ranges),
  starts=start(up.ranges)-1,
  ends=end(up.ranges),
  strands=strand(up.ranges))

write.table(up.ranges.df, file=paste0(output, "up.bed"), quote=F, sep="\t", row.names=F, col.names=F)

# Subset for down DARs and write bed file
down <- SE[rowData(SE)$dea$PValue < pthreshold & rowData(SE)$dea$logFC < -logthreshold]
down.ranges <- granges(down)

down.ranges.df <- data.frame(seqnames=seqnames(down.ranges),
  starts=start(down.ranges)-1,
  ends=end(down.ranges),
  strands=strand(down.ranges))

write.table(down.ranges.df, file=paste0(output, "down.bed"), quote=F, sep="\t", row.names=F, col.names=F)
```
Plot heatmap at DARs
```{bash engine.opts='-l', eval=F}
LiverATAC/scripts/tracks.sh LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/ctrMerged.bam LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/ logs/26_mergedtracks/ 5 8

LiverATAC/scripts/tracks.sh LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/lpdMerged.bam LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/ logs/26_mergedtracks/ 5 8

computeMatrix reference-point -R LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/up.bed LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/down.bed --referencePoint "center" -S LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/ctrMerged.bw LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/lpdMerged.bw -b 2000 -a 2000 --skipZeros -p 8 -o LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.mat.gz --outFileSortedRegions LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.bed

plotHeatmap -m LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --colorMap 'Blues_r' --whatToShow 'plot, heatmap and colorbar' --plotType se --zMin 0 --zMax 0.7 -out LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.heatmap.pdf
```



Motif analysis B6 F0 Females using Homer findMotifsGenome.pl
From the results, the enrichment scores and pvalues are manually entered into an excel sheet which is then used for plotting
```{bash engine.opts='-l', eval=F}
bed2pos.pl LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/up.bed > LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/up.pos

findMotifsGenome.pl LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/up.pos mm10 LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/up -preparsedDir LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/preparsed/ -size 200 -mask -p 8


bed2pos.pl LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/down.bed > LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/down.pos

findMotifsGenome.pl LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/down.pos mm10 LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/down -preparsedDir LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/preparsed/ -size 200 -mask -p 8
```

Load in excel sheet with enrichment scores and pvalues and plot heatmap
```{r}
motif_data <- as.data.frame(read_excel("LiverATAC/results/TFs.motifs.xlsx", sheet = 4))
rownames(motif_data) <- motif_data[,1]
motif_data <- motif_data[,-1]

# Plot using pheatmap
library(pheatmap)
    
heatmap <- pheatmap(t(motif_data), cluster_cols = F, cluster_rows = F, show_rownames = T, show_colnames = T,
                       color = colorRampPalette(c("white", "#b563f0"))(50),
                       silent = F, gaps_row = c(1:8), gaps_col = c(1:3))

pdf("LiverATAC/results/F1.motif.heatmap.pdf", width = 4, height = 3)
heatmap
dev.off()
```

Supplements


Annotate peaks
```{r}
library(ChIPseeker)

#Load in TxDB and function to clean up annotation dataframe
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
cleanupanno <- function(annostat) {

  rownames(annostat) <- annostat$Feature
  annostat$Feature <- NULL
  annostat <- as.data.frame(t(annostat))
  annostat$Exon <- sum(annostat$`1st Exon`, annostat$`Other Exon`)
  annostat$Intron <- sum(annostat$`1st Intron`, annostat$`Other Intron`)

  annostat$`1st Intron` <- NULL
  annostat$`Other Intron` <- NULL
  annostat$`1st Exon` <- NULL
  annostat$`Other Intron` <- NULL

  annostat <- as.data.frame(t(annostat))
  annostat$Feature <- rownames(annostat)
  
  annostat$FrequencyPercent <- annostat$Frequency/100
  
  return(annostat)
  
}

# Get everything for Down DARs
anno.down <- annotatePeak(down.ranges, TxDb = txdb,genomicAnnotationPriority = c("Promoter", "5UTR", "3UTR", "Exon", "Intron","Intergenic" ),
                     tssRegion = c(-1000, 1000))
annostat.down <- as.data.frame(anno.down@annoStat)
annostat.down <- cleanupanno(annostat = annostat.down)

annostat.down$DARs <- "Down DARs"

# Get everything for Up DARs
anno.up <- annotatePeak(up.ranges, TxDb = txdb,genomicAnnotationPriority = c("Promoter", "5UTR", "3UTR", "Exon", "Intron","Intergenic" ),
                     tssRegion = c(-1000, 1000))
annostat.up <- as.data.frame(anno.up@annoStat)
annostat.up <- cleanupanno(annostat = annostat.up)

annostat.up$DARs <- "Up DARs"

annostat.merged <- rbind(annostat.up, annostat.down)

plot <- ggplot(annostat.merged, aes(fill=Feature, y=Frequency, DARs)) + 
  geom_bar(position="fill", stat="identity") +
  scale_color_viridis(discrete=TRUE, option="cividis") + 
  theme_classic()

ggsave(paste0(output, "DAR.genomic.distribution.pdf"), plot, width = 6, height = 6)

```

Plot histone PTMs from public datasets at DARs
```{bash engine.opts='-l', eval=F}
computeMatrix reference-point -R LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/up.bed LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/down.bed --referencePoint "center" -S Public_data/Liver/08_tracks/Liver_H3K4me1.merged.bw Public_data/Liver/08_tracks/Liver_H3K4me3.merged.bw Public_data/Liver/08_tracks/Liver_H3K27ac.merged.bw -b 2000 -a 2000 --skipZeros -p 8 -o LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.PTMs.mat.gz --outFileSortedRegions LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.PTMs.bed

plotHeatmap -m LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.PTMs.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --heatmapHeight 4 --heatmapWidth 2 --colorMap 'GnBu_r' 'Greens_r' 'hot_r' --whatToShow 'plot, heatmap and colorbar' --plotType se --regionsLabel "H3K4me1" "H3K4me3" "H3K27ac" -out LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.PTMs.heatmap.pdf
```

Plot ATAC signal of Females in male DARs
```{bash engine.opts='-l', eval=F}
computeMatrix reference-point -R LiverATAC/results/dea_merged_07/B6-F1-Male-Basemodel_downsampled/up.bed LiverATAC/results/dea_merged_07/B6-F1-Male-Basemodel_downsampled/down.bed --referencePoint "center" -S LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/ctrMerged.bw LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/lpdMerged.bw -b 2000 -a 2000 --skipZeros -p 8 -o LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.mat.gz --outFileSortedRegions LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.bed

plotHeatmap -m LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.mat.gz --refPointLabel "Peak center" --plotFileFormat "pdf" --colorMap 'Blues_r' --whatToShow 'plot, heatmap and colorbar' --plotType se --zMin 0 --zMax 0.7 -out LiverATAC/results/dea_merged_07/B6-F1-Female-Basemodel_downsampled/DAR.heatmap.pdf
```

