---
title: "F1 analysis"
author: "Leonard C. Steg"
date: '2025-10-16'
output: html_document
---
```{r}
suppressPackageStartupMessages({
  library(edgeR)
  library(RColorBrewer)
  library(ggplot2)
  library(SummarizedExperiment)
  library(cowplot)
  library(SEtools)
  library(gprofiler2)
  library(rtracklayer)
  library(dplyr)
})
```


Read in pheno file with metadata
```{r}
pheno <- read.csv("Steg2026/LiverRNA/pheno.csv", row.names = 1)
```

Load DEA function
```{r}
source("Steg2026/LiverRNA/scripts/dea.RNA.Liver.F1.R")
```

Run DAE for male offspring of B6 fathers
```{r}
dea.RNA.Liver.F1(pheno = pheno, strain = "B6", generation = "F1", sex = "Male", mergeBy = "Father",
                 VarExp = "Group", CtrGroup = "CTR", ExpGroup = "LPD", 
                 outputdir = "Steg2026/LiverRNA/results/03_dea/", 
                 salmondir = "Steg2026/LiverRNA/results/preprocess_01/05_salmon_quant/",
                 suffix ="MergedFather_Basemodel")
```


Run DEA for female offspring of B6 fathers
```{r}
dea.RNA.Liver.F1(pheno = pheno, strain = "B6", generation = "F1", sex = "Female", mergeBy = "Father",
                 VarExp = "Group", CtrGroup = "CTR", ExpGroup = "LPD", 
                 outputdir = "Steg2026/LiverRNA/results/03_dea/", 
                 salmondir = "Steg2026/LiverRNA/results/preprocess_01/05_salmon_quant/",
                 suffix ="MergedFather_Basemodel")
```

Load functions for plotting
```{r}
source("Steg2026/scripts/VolcanoPlotGGPlot.R")

source("Steg2026/scripts/GoTermBarplot.R")
```

Load results plot volcano plot and GO bar plot for male offspring of B6 fathers
```{r}
dir <- "Steg2026/LiverRNA/results/03_dea/B6-F1-Male-MergedFather_Basemodel/"

se.males <- readRDS(paste0(dir, "SE.rds"))

VolcanoPlotGGPlot(dataframe = as.data.frame(rowData(se.males)), logFC = "dea.logFC", PValues = "dea.padj",
                  xlimits = 5, ylimits = 50,
                  color.highlight ="#3849A9" ,output = dir)

GoTermBarplot(GO.res = metadata(se.males)$go_merged, color.low = "#cbd0ee", color.high = "#3849A9",xlimits = 9,
              output = dir,print = T)


res.males <- as.data.frame(rowData(se.males))

sum(res.males$dea.padj < 0.05 & res.males$dea.logFC > 0.5)
# 38
sum(res.males$dea.padj < 0.05 & res.males$dea.logFC < -0.5)
# 13
```

Load results plot volcano plot and GO bar plot for female offspring of B6 fathers
```{r}
dir <- "Steg2026/LiverRNA/results/03_dea/B6-F1-Female-MergedFather_Basemodel/"

se.females <- readRDS(paste0(dir, "SE.rds"))

VolcanoPlotGGPlot(dataframe = as.data.frame(rowData(se.females)), logFC = "dea.logFC", PValues = "dea.padj",
                  xlimits = 5, ylimits = 50,
                  color.highlight ="#A93849" ,output = dir)

GoTermBarplot(GO.res = metadata(se.females)$go_merged, color.low = "#eecbd0", color.high = "#A93849",xlimits = 9,
              output = dir,print = T)


res.females <- as.data.frame(rowData(se.females))

sum(res.females$dea.padj < 0.05 & res.females$dea.logFC > 0.5)
# 20
sum(res.females$dea.padj < 0.05 & res.females$dea.logFC < -0.5)
# 21
```



Supplements

Violin plots for selected genes
```{r}
se.males.plot <- readRDS("Steg2026/LiverRNA/results/03_dea/B6-F1-Male-Basemodel/SE.rds")

se.females.plot <- readRDS("Steg2026/LiverRNA/results/03_dea/B6-F1-Female-Basemodel/SE.rds")

source("Steg2026/scripts/plot_violin_gene.R")

genes <- c("Pnpla3","Cyp2b9")

for (i in 1:length(genes)) {
    plot_violin_gene(se = se.males.plot, assay = "logcpm",  genes[i],
                   name = paste0("ViolinPlot_", genes[i],".pdf"), 
                   outdir = "Steg2026/LiverRNA/results/03_dea/B6-F1-Male-Basemodel/ViolinPlots/")
  
}

genes <- c("Pnpla3","Cyp2b9", "Cyp3a16")

for (i in 1:length(genes)) {
    plot_violin_gene(se = se.females.plot, assay = "logcpm",  genes[i],
                   name = paste0("ViolinPlot_", genes[i],".pdf"), 
                   outdir = "Steg2026/LiverRNA/results/03_dea/B6-F1-Female-Basemodel/ViolinPlots/")
  
}
```

Plot overlap between fathers and offspring males DEGs
```{r}
# Load results from B6 fathers
se.fathers <- readRDS("Steg2026/LiverRNA/results/03_dea/B6-F0-Male-Basemodel/SE.rds")
res.fathers <- as.data.frame(rowData(se.fathers))

males.up   <- res.males[res.males$dea.padj < 0.05 & res.males$dea.logFC >  0.5,]
males.down <- res.males[res.males$dea.padj < 0.05 & res.males$dea.logFC < -0.5,]

fathers.up   <- res.fathers[res.fathers$dea.padj < 0.05 & res.fathers$dea.logFC >  0.5,]
fathers.down <- res.fathers[res.fathers$dea.padj < 0.05 & res.fathers$dea.logFC < -0.5,]

overlap.up   <- males.up[rownames(males.up) %in% rownames(fathers.up),]
overlap.down <- males.down[rownames(males.down) %in% rownames(fathers.down),]

# Plot as barplot (counts)
df <- as.data.frame(matrix(NA, nrow = 6, ncol = 3,
                           dimnames = list(NULL, c("Direction", "Overlap", "Number"))))

df$Direction <- c(rep("Up", 3), rep("Down", 3))
df$Overlap   <- rep(c("Male offspring specific", "Overlap", "Father specific"), 2)

df[1, "Number"] <- nrow(males.up)     - nrow(overlap.up)
df[2, "Number"] <- nrow(overlap.up)
df[3, "Number"] <- nrow(fathers.up)   - nrow(overlap.up)

df[4, "Number"] <- nrow(males.down)   - nrow(overlap.down)
df[5, "Number"] <- nrow(overlap.down)
df[6, "Number"] <- nrow(fathers.down) - nrow(overlap.down)

# Order factors: consistent stack + Down left, Up right
df$Overlap <- factor(df$Overlap, levels = c("Father specific", "Overlap", "Male offspring specific"))
df$Direction <- factor(df$Direction, levels = c("Down", "Up"))

# Totals for label above bars
totals <- df %>%
  group_by(Direction) %>%
  summarise(Total = sum(Number), .groups = "drop")

overlap.fathers.males <- ggplot(df, aes(x = Direction, y = Number, fill = Overlap)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = ifelse(Number > 0, Number, "")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  geom_text(
    data = totals,
    aes(x = Direction, y = Total, label = Total),
    inherit.aes = FALSE,
    vjust = -0.3,
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Father specific"         = "#404040",
      "Overlap"                 = "#666666",
      "Male offspring specific" = "#3849A9"
    )
  ) +
  labs(
    x = NULL,
    y = "Number of DEGs",
    fill = "Expression",
    title = "Overlaps between father and male offspring liver RNA-seq results"
  ) +
  ylim(c(0,150)) + 
  theme_cowplot(font_size = 8) +
  coord_cartesian(clip = "off")

ggsave(
  "overlap.fathers.males.number.pdf",
  plot = overlap.fathers.males,
  device = NULL,
  path = "Steg2026/LiverRNA/results/",
  width = 6, height = 5, units = "cm"
)

```


Plot overlap between male and female offspring DEGs
```{r}

males.up   <- res.males[res.males$dea.padj < 0.05 & res.males$dea.logFC >  0.5,]
males.down <- res.males[res.males$dea.padj < 0.05 & res.males$dea.logFC < -0.5,]

females.up   <- res.females[res.females$dea.padj < 0.05 & res.females$dea.logFC >  0.5,]
females.down <- res.females[res.females$dea.padj < 0.05 & res.females$dea.logFC < -0.5,]

overlap.up   <- males.up[rownames(males.up) %in% rownames(females.up),]
overlap.down <- males.down[rownames(males.down) %in% rownames(females.down),]

# Plot as barplot (counts)
df <- as.data.frame(matrix(NA, nrow = 6, ncol = 3,
                           dimnames = list(NULL, c("Direction", "Overlap", "Number"))))

df$Direction <- c(rep("Up", 3), rep("Down", 3))
df$Overlap   <- rep(c("Male offspring specific", "Overlap", "Female offspring specific"), 2)

df[1, "Number"] <- nrow(males.up)     - nrow(overlap.up)
df[2, "Number"] <- nrow(overlap.up)
df[3, "Number"] <- nrow(females.up)   - nrow(overlap.up)

df[4, "Number"] <- nrow(males.down)   - nrow(overlap.down)
df[5, "Number"] <- nrow(overlap.down)
df[6, "Number"] <- nrow(females.down) - nrow(overlap.down)

# Order factors: consistent stack + Down left, Up right
df$Overlap <- factor(df$Overlap, levels = c("Male offspring specific", "Overlap", "Female offspring specific"))
df$Direction <- factor(df$Direction, levels = c("Down", "Up"))

# Totals for label above bars
totals <- df %>%
  group_by(Direction) %>%
  summarise(Total = sum(Number), .groups = "drop")

overlap.females.males <- ggplot(df, aes(x = Direction, y = Number, fill = Overlap)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = ifelse(Number > 0, Number, "")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  geom_text(
    data = totals,
    aes(x = Direction, y = Total, label = Total),
    inherit.aes = FALSE,
    vjust = -0.3,
    size = 3
  ) +
  scale_fill_manual(
    values = c(
      "Male offspring specific"   = "#3849A9",
      "Overlap"                   = "#666666",
      "Female offspring specific" = "#A93849"
    )
  ) +
  labs(
    x = NULL,
    y = "Number of DEGs",
    fill = "Expression",
    title = "Overlaps between male and female offspring liver RNA-seq results"
  ) +
  ylim(c(0,150)) + 
  theme_cowplot(font_size = 8) +
  coord_cartesian(clip = "off")

ggsave(
  "overlap.females.males.offspring.pdf",
  plot = overlap.females.males,
  device = NULL,
  path = "Steg2026/LiverRNA/results/",
  width = 6, height = 5, units = "cm"
)

```

Prepare a matrix with the allelic expression of each gene in sample

Load in maternal and paternal count tables
```{r}
# Load in maternal and paternal count tables

# Read in counts from salmon using tximport
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
# Extract all the transcript names (TXNAME) as keys
k <- keys(txdb, keytype = "TXNAME")
# Create a mapping (tx2gene) from TXNAME (transcript IDs) to GENEID (UCSC Entrez IDs)
tx2gene <- select(txdb, keys = k, columns = "GENEID", keytype = "TXNAME")
# Use org.Mm.eg.db to map UCSC Entrez IDs to Ensembl Gene IDs
ensembl_map <- mapIds(org.Mm.eg.db, 
                      keys = tx2gene$GENEID, 
                      column = "ENSEMBL", 
                      keytype = "ENTREZID", 
                      multiVals = "first")  # Takes only the first value if multiple mappings exist
# Add the Ensembl gene IDs to the tx2gene data frame
tx2gene$ENSEMBL <- ensembl_map[tx2gene$GENEID]
tx2gene_mapped <- tx2gene[!is.na(tx2gene$GENEID) & !is.na(tx2gene$ENSEMBL), ]
# Convert the ENSEMBL column from a list to a character vector
tx2gene_mapped$ENSEMBL <- sapply(tx2gene_mapped$ENSEMBL, `[`, 1)


# Define the directories with Salmon quant.sf files
samples <- rownames(pheno)
sample_dirs <- c(paste0(samples, "_genome1_quant"),paste0(samples, "_genome2_quant"))
# Manually map the sample names to the actual directory names with _SX suffix 
files <- file.path("Steg2026/LiverRNA/results/preprocess_01/12_salmon_post_snpsplit", sample_dirs, "quant.sf")
names(files) <- c(paste0(samples, "_g1"), paste0(samples, "_g2"))

if (!all(file.exists(files))){
  stop("Not all salmon files detected, please check salmondir argument.")
}

# Use tximport and extract counts
txi <- tximport(files, type = "salmon", tx2gene = tx2gene_mapped[, c("TXNAME", "ENSEMBL")])
counts <- txi$counts


# Split into maternal paternal counts
# Genome1 = B6, Genome2 = PWK
# Paternal genomes: B6_g1 and PWK_g2
# Maternal genomes: B6_g2 and PWK_g1

counts_pat <- counts[,c(grep("^B6.*g1$", names(files), value = TRUE),grep("^PWK.*g2$", names(files), value = TRUE))]

counts_mat <- counts[,c(grep("^B6.*g2$", names(files), value = TRUE),grep("^PWK.*g1$", names(files), value = TRUE))]

identical(gsub(".{3}$", "", colnames(counts_mat)),gsub(".{3}$", "", colnames(counts_pat)))

# Calculate the percentage of maternal expressed transcripts for each gene and sample
perc_mat <- as.data.frame(matrix(nrow = nrow(counts), ncol = ncol(counts_mat), dimnames = dimnames(counts_mat)))

colnames(perc_mat) <- gsub(".{3}$", "", colnames(perc_mat))

for (i in 1:nrow(perc_mat)){
  for (j in 1:ncol(perc_mat)){
    perc_mat[i,j] <- counts_mat[i,j]/(counts_mat[i,j]+counts_pat[i,j])
    
  }
  
}



# Filtering the matrix to remove columns that have non-useful numbers.

# Apply filterbyExpr on count matrix with groups being a vector of "Maternal" or "Paternal"
group <- rep("Mat", ncol(counts))
group[colnames(counts) %in% colnames(counts_pat)] <- "Pat"

keep_filterbyExpr <- filterByExpr(counts <- counts, group = group)
perc_mat <- perc_mat[keep_filterbyExpr,]

# Add gene names instead of ENSG identifiers

# Use org.Mm.eg.db to map Ensembl Gene IDs to Symbols
rownames(perc_mat) <- mapIds(org.Mm.eg.db, 
                      keys = rownames(perc_mat), 
                      column = "SYMBOL", 
                      keytype = "ENSEMBL", 
                      multiVals = "first") 

# Save as Rds
saveRDS(perc_mat, file = "Steg2026/LiverRNA/results/perc_mat.Rds")
```


Allele specific analysis 1: Proportions of DEGs expressed from both alleles or from maternal/paternal alleles specifically

In male offspring
```{r}
males <- pheno[pheno$Generation == "F1" & pheno$Strain == "B6" & pheno$Sex == "Male",]$Sample

perc.mat.males <- perc.mat[,males]

# Calculate the median per gene
row_medians <- apply(perc.mat.males, 1, median)

# Convert the named numeric vector to a data frame
median_df <- data.frame(
  name = names(row_medians),
  row_median = as.numeric(row_medians)
)

# How many of the male DEGs are expressed on the paternal allele (median < 0.2), on both alleles (median >0.2 <0.8) or on maternal allele (median > 0.8)

res.males <-   left_join(res.males, median_df, by = "name")

# Plot stacked barplots of allele information of DEGs
plot_df <- res.males %>%
  filter(!is.na(dea.padj), !is.na(dea.logFC)) %>%
  mutate(
    direction = case_when(
      dea.padj < 0.05 & dea.logFC < -0.5 ~ "Downregulated",
      dea.padj < 0.05 & dea.logFC > 0.5  ~ "Upregulated",
      TRUE ~ NA_character_
    ),
    expr_class = case_when(
      is.na(row_median)            ~ "No allele information",
      row_median < 0.2             ~ "Paternally expressed",
      row_median > 0.8             ~ "Maternally expressed",
      row_median >= 0.2 & row_median <= 0.8 ~ "Bi-allelic",
      TRUE ~ "No allele information"
    )
  ) %>%
  filter(!is.na(direction))

# Summarize to counts + percentages
plot_summ <- plot_df %>%
  count(direction, expr_class, name = "n") %>%
  group_by(direction) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(
    direction = factor(direction, levels = c("Downregulated", "Upregulated")),
    expr_class = fct_relevel(expr_class,
                             "Paternally expressed",
                             "Bi-allelic",
                             "Maternally expressed",
                             "No allele information")
  )

# Plot
allelic.barplot.males <- ggplot(plot_summ, aes(x = direction, y = pct, fill = expr_class)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = scales::percent(pct, accuracy = 1)),
            position = position_stack(vjust = 0.5),
            size = 4) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Percentage of DEGs",
       fill = "Expression class",
       title = "DEG by allelic expression") +
  theme_cowplot(font_size = 8)


ggsave(filename =  "Steg2026/LiverRNA/results/03_dea/B6-F1-Male-MergedFather_Basemodel/allelic.barplot.males.pdf", allelic.barplot.males,  width = 8, height = 5, units = "cm")

```

In female offspring
```{r}
females <- pheno[pheno$Generation == "F1" & pheno$Strain == "B6" & pheno$Sex == "Female",]$Sample

perc.mat.females <- perc.mat[,females]

# Calculate the median per gene
row_medians <- apply(perc.mat.females, 1, median)

# Convert the named numeric vector to a data frame
median_df <- data.frame(
  name = names(row_medians),
  row_median = as.numeric(row_medians)
)

# How many of the male DEGs are expressed on the paternal allele (median < 0.2), on both alleles (median >0.2 <0.8) or on maternal allele (median > 0.8)
res.females <-   left_join(res.females, median_df, by = "name")

# Plot stacked barplots of allele information of DEGs
plot_df <- res.females %>%
  filter(!is.na(dea.padj), !is.na(dea.logFC)) %>%
  mutate(
    direction = case_when(
      dea.padj < 0.05 & dea.logFC < -0.5 ~ "Downregulated",
      dea.padj < 0.05 & dea.logFC > 0.5  ~ "Upregulated",
      TRUE ~ NA_character_
    ),
    expr_class = case_when(
      is.na(row_median)            ~ "No allele information",
      row_median < 0.2             ~ "Paternally expressed",
      row_median > 0.8             ~ "Maternally expressed",
      row_median >= 0.2 & row_median <= 0.8 ~ "Bi-allelic",
      TRUE ~ "No allele information"
    )
  ) %>%
  filter(!is.na(direction))

# Summarize to counts + percentages
plot_summ <- plot_df %>%
  count(direction, expr_class, name = "n") %>%
  group_by(direction) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(
    direction = factor(direction, levels = c("Downregulated", "Upregulated")),
    expr_class = fct_relevel(expr_class,
                             "Paternally expressed",
                             "Bi-allelic",
                             "Maternally expressed",
                             "No allele information")
  )

# Plot
allelic.barplot.females <- ggplot(plot_summ, aes(x = direction, y = pct, fill = expr_class)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = scales::percent(pct, accuracy = 1)),
            position = position_stack(vjust = 0.5),
            size = 4) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Percentage of DEGs",
       fill = "Expression class",
       title = "DEGs by allelic expression") +
  theme_cowplot(font_size = 8)
  
ggsave(filename =  "Steg2026/LiverRNA/results/03_dea/B6-F1-Female-MergedFather_Basemodel/allelic.barplot.females.pdf", allelic.barplot.females,  width = 8, height = 5, units = "cm")
```


Allele specific analysis 2: Test whether DEGs change their allelic expression patterns due to paternal LPD

In male offspring
```{r}
# Prepare dataframe
pheno.males <- pheno[pheno$Generation == "F1" & pheno$Strain == "B6" & 
                       pheno$Sex == "Male",]

perc.mat.males <- perc.mat.males[,pheno.males$Sample]

perc.mat.males <- perc.mat.males[res.males[res.males$dea.padj < 0.05 & abs(res.males$dea.logFC) > 0.5,]$name,]

perc.mat.males <- perc.mat.males[complete.cases(perc.mat.males),]

allele.switch.males <- as.data.frame(matrix(NA, nrow = nrow(perc.mat.males), ncol = 4, 
                                            dimnames = list(rownames(perc.mat.males), 
                                                            c("PValue", "Mean LPD", "Mean CTR", "Diff"))))


for (i in 1:nrow(allele.switch.males)){
  
  test.results <- t.test(as.numeric(perc.mat.males[i,]) ~ as.factor(pheno.males$Group))
  
  allele.switch.males[i,]$PValue <- test.results$p.value
  allele.switch.males[i,]$`Mean LPD` <- test.results$estimate[1]
  allele.switch.males[i,]$`Mean CTR` <- test.results$estimate[2]
  allele.switch.males[i,]$Diff <- allele.switch.males[i,]$`Mean LPD`-allele.switch.males[i,]$`Mean CTR`
}

allele.switch.males$padj <- p.adjust(allele.switch.males$PValue, method = "BH")

group_df <- pheno.males[,c("Sample", "Group")]

# Choose the genes to plot 
genes_to_plot <- rownames(perc.mat.males)

# Convert to long format + join group labels
df_long <- perc.mat.males[genes_to_plot, , drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("Gene") %>%
  pivot_longer(-Gene, names_to = "Sample", values_to = "Perc.Mat") %>%
  left_join(group_df, by = "Sample") %>%
  mutate(
    # order genes as they appear in genes_to_plot, shown top-to-bottom
    Gene  = factor(Gene, levels = rev(genes_to_plot)),
    # ensure Control appears before LPD in the legend & dodge
    Group = factor(Group, levels = c("CTR", "LPD"))
  )

# Order genes by median Perc.Mat (across all samples)
gene_order <- df_long %>%
  group_by(Gene) %>%
  summarise(median_value = median(Perc.Mat, na.rm = TRUE)) %>%
  arrange(desc(median_value)) %>%
  pull(Gene)

df_long <- df_long %>%
  mutate(
    Gene = factor(Gene, levels = gene_order),
    Group = factor(Group, levels = c("CTR", "LPD"))
  )

# Plot and save
allele.switch.DEGs.males.plot <- ggplot(df_long, aes(x = Gene, y = Perc.Mat, fill = Group)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, outlier.size = 1.4) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Perc.Mat",
    title = "Perc.Mat across genes by Group"
  ) +
  theme_cowplot(font_size = 8, line_size = 0.3) +
  theme(
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) 

ggsave(filename =  "Steg2026/LiverRNA/results/03_dea/B6-F1-Male-MergedFather_Basemodel/allele.switch.DEGs.males.pdf", allele.switch.DEGs.males.plot,  width = 8, height = 15, units = "cm")

```

In female offspring
```{r}
# Prepare dataframe
pheno.females <- pheno[pheno$Generation == "F1" & pheno$Strain == "B6" & 
                       pheno$Sex == "Female",]

perc.mat.females <- perc.mat.females[,pheno.females$Sample]

perc.mat.females <- perc.mat.females[res.females[res.females$dea.padj < 0.05 & abs(res.females$dea.logFC) > 0.5,]$name,]

perc.mat.females <- perc.mat.females[complete.cases(perc.mat.females),]

allele.switch.females <- as.data.frame(matrix(NA, nrow = nrow(perc.mat.females), ncol = 4, 
                                            dimnames = list(rownames(perc.mat.females), 
                                                            c("PValue", "Mean LPD", "Mean CTR", "Diff"))))


for (i in 1:nrow(allele.switch.females)){
  
  test.results <- t.test(as.numeric(perc.mat.females[i,]) ~ pheno.females$Group)
  
  allele.switch.females[i,]$PValue <- test.results$p.value
  allele.switch.females[i,]$`Mean LPD` <- test.results$estimate[1]
  allele.switch.females[i,]$`Mean CTR` <- test.results$estimate[2]
  allele.switch.females[i,]$Diff <- allele.switch.females[i,]$`Mean LPD`-allele.switch.females[i,]$`Mean CTR`
}

allele.switch.females$padj <- p.adjust(allele.switch.females$PValue, method = "BH")


group_df <- pheno.females[,c("Sample", "Group")]

# Choose the genes to plot 
genes_to_plot <- rownames(perc.mat.females)

# Convert to long format + join group labels
df_long <- perc.mat.females[genes_to_plot, , drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("Gene") %>%
  pivot_longer(-Gene, names_to = "Sample", values_to = "Perc.Mat") %>%
  left_join(group_df, by = "Sample") %>%
  mutate(
    # order genes as they appear in genes_to_plot, shown top-to-bottom
    Gene  = factor(Gene, levels = rev(genes_to_plot)),
    # ensure Control appears before LPD in the legend & dodge
    Group = factor(Group, levels = c("CTR", "LPD"))
  )

#Order genes by median Perc.Mat (across all samples)
gene_order <- df_long %>%
  group_by(Gene) %>%
  summarise(median_value = median(Perc.Mat, na.rm = TRUE)) %>%
  arrange(desc(median_value)) %>%
  pull(Gene)

df_long <- df_long %>%
  mutate(
    Gene = factor(Gene, levels = gene_order),
    Group = factor(Group, levels = c("CTR", "LPD"))
  )

# Plot and save
allele.switch.DEGs.females.plot <- ggplot(df_long, aes(x = Gene, y = Perc.Mat, fill = Group)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, outlier.size = 1.4) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Perc.Mat",
    title = "Perc.Mat across genes by Group"
  ) +
  theme_cowplot(font_size = 8, line_size = 0.3) +
  theme(
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) 

ggsave(filename =  "Steg2026/LiverRNA/results/03_dea/B6-F1-Female-MergedFather_Basemodel/allele.switch.DEGs.females.pdf", allele.switch.DEGs.females.plot,  width = 8, height = 15, units = "cm")
```
